Extracellular Fluxes

### 2022-06-06 analysis/extracellular-fluxes.Rmd
add q bias correction analysis

```{r}
data_raw %>% 
  select(cell_type, experiment, date) %>% 
  distinct() %>% 
  group_by(cell_type, experiment) %>% 
  summarise(count = n()) %>% 
  my_kable(
    col.names = c("Cell type", "Experiment", "Count"), 
    caption = "Biological replicates"
  )
```

@@

std %>% 
  select(metabolite:run, summary) %>% 
  unnest(c(summary)) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(
    title = str_c(cell_type, experiment, sep = " "), 
    plots = map2(data, title, plot_r_squared)
  ) %>% 
  pull(plots)
```

@@

```{r, results = "hide", fig.show = "hold", fig.align = "left", out.width = "24.5%"}
std %>% 
  filter(detector == "fld", date == "2018-02-12", experiment == "02") %>% 
  pull(plots)
```

@@

```{r, fig.show = "hold", fig.align = "left", results = "hide"}
std_fld %>% 
  select(metabolite:run, summary) %>% 
  unnest(c(summary)) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(
    title = str_c(cell_type, experiment, sep = " "), 
    plots = map2(data, title, plot_r_squared)
  ) %>% 
  pull(plots)
```

@@

```{r}
poor_fits <- 
  std_fld %>% 
  ungroup() %>% 
  select(metabolite:run, summary, plots) %>% 
  unnest(c(summary)) %>% 
  slice_min(order_by = r.squared, n = 7)

poor_fits %>% 
  select(metabolite:r.squared) %>% 
  my_kable(caption = "Poor standard curve fits")
```

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
poor_fits %>% 
  pull(plots)
```

@@

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
std_clean %>% 
  semi_join(poor_fits, by = c("metabolite", "experiment", "date", "batch", "run", "detector")) %>% 
  pull(plots)
```

@@

```{r}
pyruvate <- 
  conc %>% 
  filter(metabolite == "pyruvate") %>%
  mutate(detector = factor(
    detector, 
    levels = c("enzyme", "hplc", "lcms"), 
    labels = c("Enzyme", "HPLC", "LC-MS")
  ))

pyruvate %>% 
  filter(conc < 1500) %>% 
  filter(time == 0) %>% 
  ggplot() + 
  aes(x = detector, y = conc, color = detector) +
  geom_jitter(show.legend = FALSE) +
  labs(
    title = "Pyruvate concentrations at time 0", 
    x = "Detection method", 
    y = "Concentration (μM)"
  )
```

@@

```{r, fig.show = "hold"}
amino_acids <-
  conc %>% 
  filter(detector %in% c("mwd", "fld"))

plot_aa <- function(df) {
  df %>% 
    filter(!is.na(conc)) %>% 
    ggplot() + 
    aes(y = conc, x = metabolite, color = detector) + 
    geom_hline(yintercept = 300, linetype = 2, color = "gray") +
    geom_jitter(alpha = 0.1) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    labs(
      x = "Amino acid", 
      y = "Concentration (μM)"
    ) + 
    scale_color_discrete(name = "Detector", labels = c("FLD", "MWD")) +
    guides(color = guide_legend(override.aes = list(alpha = 1)))
}

amino_acids %>% 
  filter(cell_type == "lf") %>% 
  plot_aa() + ggtitle("LF amino acids")

amino_acids %>% 
  filter(cell_type == "pasmc") %>% 
  plot_aa() + ggtitle("PASMC amino acids")
```

@@

```{r}
amino_acids %>% 
  group_by(metabolite) %>% 
  summarise(conc = mean(conc, na.rm = TRUE)) %>% 
  mutate(detector = case_when(
    metabolite == "cystine" ~ "mwd", 
    conc > 300 ~ "mwd", 
    conc <= 300 ~ "fld"
  )) %>% 
  filter(detector == "mwd") %>% 
  select(metabolite) %>% 
  my_kable(caption = "MWD detection", col.names = "Metabolite")
```

@@

```{r}
evap %>% 
  ggplot() +
  aes(
    x = time, 
    y = volume, 
    color = interaction(oxygen, experiment, sep = " | ")
  ) +
  geom_point(
    alpha = 0.3, 
    size = 2
  ) +
  geom_smooth(
    method = "lm", 
    formula = y ~ x, 
    se = FALSE, 
    show.legend = FALSE
  ) +
  stat_summary(
    geom = "pointrange", 
    fun.data = "mean_se", 
    alpha = 1, 
    fatten = 1, 
    show.legend = FALSE
  ) +
  guides(color = guide_legend(
    title = "Treatment", 
    override.aes = list(alpha = 1)
  )) +
  labs(
    tite = "Predicted well volumes", 
    x = "Time (h)", 
    y = "Volume (ml)"
  )
```

@@

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
growth_curves %>% 
  select(cell_type:data) %>% 
  unnest(c(data)) %>% 
  group_by(cell_type, experiment, batch, date, oxygen, virus, treatment, time) %>%
  summarize(conc = mean(conc, na.rm = TRUE)) %>% 
  filter(!is.na(time)) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(plot = map2(
    data, 
    cell_type, 
    ~ggplot(.x) + 
      aes(
        x = time, 
        y = conc/1000, 
        color = interaction(oxygen, treatment, virus, sep = " | ")
      ) +
      stat_summary(
        geom = "pointrange", 
        fun.data = "mean_se"
      ) +
      stat_summary(
        geom = "line", 
        fun.data = "mean_se", 
        show.legend = FALSE
      ) +
      labs(
        title = toupper(.y), 
        x = "Time (h)", 
        y = expression(paste("Cell count (\U00D7", "10"^3, ")")), 
        color = "condition"
      )
  )) %>% 
  pull(plot)
```

@@

fluxes %>%
  filter(metabolite %in% c("glucose", "lactate")) %>%
  plot_fluxes("Metabolite flux") + facet_wrap(~cell_type, nrow = 1)
```

```{r, results = "hide", out.width = "100%"}
fluxes %>% 
  filter(!(metabolite %in% c("glucose", "lactate"))) %>% 
  group_by(cell_type, experiment) %>% 
  nest() %>% 
  mutate(plots = map2(data, toupper(cell_type), plot_fluxes)) %>% 
  pull(plots)
```

@@

```{r}
fluxes %>%
  ungroup() %>%
  group_by(metabolite, cell_type, oxygen, treatment, virus) %>%
  summarise(
    v = mean(flux, na.rm = TRUE),
    sd = sd(flux, na.rm = TRUE),
    n = n(),
    se = sd / sqrt(n()),
    cv = sd / abs(v)
  ) %>%
  arrange(cell_type, metabolite, oxygen, treatment, virus) %>%
  my_kable(
    caption = "Flux rate summary",
    col.names = c(
      "Metabolite", 
      "Cell type", 
      "Oxygen", 
      "Treatment", 
      "Virus", 
      "Flux", 
      "SD", 
      "n", 
      "SEM", 
      "CV"
    ),
    align = "lccccccccc",
    digits = 2,
    longtable = TRUE,
  ) %>%
  kableExtra::kable_styling(
    latex_options = c("repeat_header"),
    font_size = 8
  )
```

>>>>

```{r}
data_raw |> 
  select(cell_type, experiment, date) |> 
  distinct() |> 
  group_by(cell_type, experiment) |> 
  summarise(count = n()) |> 
  my_kable(
    col.names = c("Cell type", "Experiment", "Count"), 
    caption = "Biological replicates"
  )
```

@@

std |> 
  select(metabolite:run, summary) |> 
  unnest(c(summary)) |> 
  group_by(cell_type, experiment) |> 
  nest() |> 
  mutate(
    title = str_c(cell_type, experiment, sep = " "), 
    plots = map2(data, title, plot_r_squared)
  ) |> 
  pull(plots)
```

@@

```{r, results = "hide", fig.show = "hold", fig.align = "left", out.width = "24.5%"}
std |> 
  filter(detector == "fld", date == "2018-02-12", experiment == "02") |> 
  pull(plots)
```

@@

```{r, fig.show = "hold", fig.align = "left", results = "hide"}
std_fld |> 
  select(metabolite:run, summary) |> 
  unnest(c(summary)) |> 
  group_by(cell_type, experiment) |> 
  nest() |> 
  mutate(
    title = str_c(cell_type, experiment, sep = " "), 
    plots = map2(data, title, plot_r_squared)
  ) |> 
  pull(plots)
```

@@

```{r}
poor_fits <- 
  std_fld |> 
  ungroup() |> 
  select(metabolite:run, summary, plots) |> 
  unnest(c(summary)) |> 
  slice_min(order_by = r.squared, n = 7)

poor_fits |> 
  select(metabolite:r.squared) |> 
  my_kable(caption = "Poor standard curve fits")
```

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
poor_fits |> 
  pull(plots)
```

@@

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
std_clean |> 
  semi_join(poor_fits, by = c("metabolite", "experiment", "date", "batch", "run", "detector")) |> 
  pull(plots)
```

@@

```{r}
pyruvate <- 
  conc |> 
  filter(metabolite == "pyruvate") |>
  mutate(detector = factor(
    detector, 
    levels = c("enzyme", "hplc", "lcms"), 
    labels = c("Enzyme", "HPLC", "LC-MS")
  ))

pyruvate |> 
  filter(conc < 1500) |> 
  filter(time == 0) |> 
  ggplot() + 
  aes(x = detector, y = conc, color = detector) +
  geom_jitter(show.legend = FALSE) +
  labs(
    title = "Pyruvate concentrations at time 0", 
    x = "Detection method", 
    y = "Concentration (μM)"
  )
```

@@

```{r, fig.show = "hold"}
amino_acids <-
  conc |> 
  filter(detector %in% c("mwd", "fld"))

plot_aa <- function(df) {
  df |> 
    filter(!is.na(conc)) |> 
    ggplot() + 
    aes(y = conc, x = metabolite, color = detector) + 
    geom_hline(yintercept = 300, linetype = 2, color = "gray") +
    geom_jitter(alpha = 0.1) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +
    labs(
      x = "Amino acid", 
      y = "Concentration (μM)"
    ) + 
    scale_color_discrete(name = "Detector", labels = c("FLD", "MWD")) +
    guides(color = guide_legend(override.aes = list(alpha = 1)))
}

amino_acids |> 
  filter(cell_type == "lf") |> 
  plot_aa() + ggtitle("LF amino acids")

amino_acids |> 
  filter(cell_type == "pasmc") |> 
  plot_aa() + ggtitle("PASMC amino acids")
```

@@

```{r}
amino_acids |> 
  group_by(metabolite) |> 
  summarise(conc = mean(conc, na.rm = TRUE)) |> 
  mutate(detector = case_when(
    metabolite == "cystine" ~ "mwd", 
    conc > 300 ~ "mwd", 
    conc <= 300 ~ "fld"
  )) |> 
  filter(detector == "mwd") |> 
  select(metabolite) |> 
  my_kable(caption = "MWD detection", col.names = "Metabolite")
```

@@

```{r}
evap |> 
  ggplot() +
  aes(
    x = time, 
    y = volume, 
    color = interaction(oxygen, experiment, sep = " | ")
  ) +
  geom_point(
    alpha = 0.3, 
    size = 2
  ) +
  geom_smooth(
    method = "lm", 
    formula = y ~ x, 
    se = FALSE, 
    show.legend = FALSE
  ) +
  stat_summary(
    geom = "pointrange", 
    fun.data = "mean_se", 
    alpha = 1, 
    fatten = 1, 
    show.legend = FALSE
  ) +
  guides(color = guide_legend(
    title = "Treatment", 
    override.aes = list(alpha = 1)
  )) +
  labs(
    tite = "Predicted well volumes", 
    x = "Time (h)", 
    y = "Volume (ml)"
  )
```

@@

```{r, fig.show = "hold", fig.align = "left", results = "hide", out.width = "32.9%"}
growth_curves |> 
  select(cell_type:data) |> 
  unnest(c(data)) |> 
  group_by(cell_type, experiment, batch, date, oxygen, virus, treatment, time) |>
  summarize(conc = mean(conc, na.rm = TRUE)) |> 
  filter(!is.na(time)) |> 
  group_by(cell_type, experiment) |> 
  nest() |> 
  mutate(plot = map2(
    data, 
    cell_type, 
    ~ggplot(.x) + 
      aes(
        x = time, 
        y = conc/1000, 
        color = interaction(oxygen, treatment, virus, sep = " | ")
      ) +
      stat_summary(
        geom = "pointrange", 
        fun.data = "mean_se"
      ) +
      stat_summary(
        geom = "line", 
        fun.data = "mean_se", 
        show.legend = FALSE
      ) +
      labs(
        title = toupper(.y), 
        x = "Time (h)", 
        y = expression(paste("Cell count (\U00D7", "10"^3, ")")), 
        color = "condition"
      )
  )) |> 
  pull(plot)
```

@@

fluxes |>
  filter(metabolite %in% c("glucose", "lactate")) |>
  plot_fluxes("Metabolite flux") + facet_wrap(~cell_type, nrow = 1)
```

```{r, results = "hide", out.width = "100%"}
fluxes |> 
  filter(!(metabolite %in% c("glucose", "lactate"))) |> 
  group_by(cell_type, experiment) |> 
  nest() |> 
  mutate(plots = map2(data, toupper(cell_type), plot_fluxes)) |> 
  pull(plots)
```

@@

```{r}
fluxes |>
  ungroup() |>
  group_by(metabolite, cell_type, oxygen, treatment, virus) |>
  summarise(
    v = mean(flux, na.rm = TRUE),
    sd = sd(flux, na.rm = TRUE),
    n = n(),
    se = sd / sqrt(n()),
    cv = sd / abs(v)
  ) |>
  arrange(cell_type, metabolite, oxygen, treatment, virus) |>
  my_kable(
    caption = "Flux rate summary",
    col.names = c(
      "Metabolite", 
      "Cell type", 
      "Oxygen", 
      "Treatment", 
      "Virus", 
      "Flux", 
      "SD", 
      "n", 
      "SEM", 
      "CV"
    ),
    align = "lccccccccc",
    digits = 2,
    longtable = TRUE,
  ) |>
  kableExtra::kable_styling(
    latex_options = c("repeat_header"),
    font_size = 8
  )
```
